---
title: Factors affecting frequency of Volcanic eruptions
author: "x404-pagenotfound"
output: github_document
---

## Summary

We chose to research what factors affect the frequency of volcanic eruptions.

The data we are analyzing comes from the tidytuesdy package, and was collected by the Smithsonian's Global Volcanism Program. To explore the factors affecting volcanic eruptions, we used 2 data frames, one of which listed every volcano on the planet, giving its position, name, number...etc, while the other data set listed every recorded volcanic eruption. Both dataframes have far too many variables to explore, so we decided to focus on factors like position, country, crust type, and tectonic setting.

Now, the easiest visualization to spot trends by location is obviously a world map, so we will start there. To get an idea of which countries have had the most volcanic eruptions, we created a choropleth map which shows number of eruptions by country. Now, this is helpful, but it is impossible to see any trend, because we can't see groups of eruptions within the countries. Also, as you can see from the scale, this map only gives a rough idea of where the most eruptions happen.

So to solve this, we create a clustered world map. Now the patterns a becoming a little more apparent. For example, we can now see that most of the large clusters of eruptions are circled around the pacific ocean. Our hypothesis as to why these trends exist had to do with tectonic plates.

So we layered a map of the tectonic borders on to the points, and as you can see, there is an clear pattern. Most of the eruptions took place along a tectonic border, which isn't surprising because volcanoes are formed by the clashing of two plates. The circle of volcanoes around the pacific that we saw earlier is a result of the pacific plate, and is actually called the Pacific Ring of Fire. So just by looking at the maps, we can see that location has a big effect on the count of eruptions in a given area.

We then wanted to investigate whether the type of crust and tectonic setting influenced the number of eruptions. To clarify these variable's meaning; the earth’s crust is split into tectonic plates which lie on top of the mantle. 
The type of plate is given as the crust variable and is either: 
-Continental (thick land plates)
-Oceanic (thinner but heavier sea plates)
-Intermediate
-And unknown plates. 
The tectonic setting is the movement of the plates and mantle that causes the volcano to form. It is given as the variables:
-Subduction zone (an oceanic plate moves beneath a continental plate)
-Rift Zone (two plates move apart)
-Intraplate (a hotter area causes increased magma at the surface)

We investigated these as when comparing the maps of distribution it was clear that certain areas of the map had higher frequencies of distinct tectonic settings and crust types in the same areas.

The initial plan was to fit linear models for frequency by crust and tectonic setting separately. However when evaluating both models, the r squared value was very low and the residuals were in distinct clusters by variable category, hence we concluded the linear models weren’t suitable for the data in hand. We decided our research question could be answered by visualization and comparisons.

When reviewing the number of eruptions on each crust type it is clear that continental crust is the crust type with the most eruptions. Comparing the number of eruptions for each tectonic setting, the subduction zone has the most eruptions. When referring back to the distribution maps this connection is shown as all of the subduction volcanoes are on land, whilst most ocean volcanos appear to be rift volcanos, a scientific study of why this is the case would be really interesting if future research occurred.

**Emma **

In conclusion we believe the factors of:
-Location of volcano (position on tectonic plate)
-Tectonic Plate Setting
-Tectonic Plate Type
-Crust Type
-Primary Volcano Type
have an influence on the number of volcano erruptions. Through our investigation we have come across many new potential relationships between these factors, which should further research be conducted with perhaps more expert knowledge on the subject could be really beneficial to looking at which areas are and could potentially become more dangerous places for living and biodiversity development.



```{r packages, echo=FALSE}
library(tidyverse)
library(broom)
library(janitor)
library(glue)
```

```{r load-data, echo = FALSE}
#setwd("/cloud/project/proposal")
volcano <- read_csv("/cloud/project/data/volcano.csv")
eruptions <- read_csv("/cloud/project/data/eruptions.csv")
events <- read_csv("/cloud/project/data/events.csv")
```

## Presentation

Our presentation can be found [here](presentation/presentation.html).

## Data

Mock, T 2020, "Volcano Eruptions", electronic dataset, Global Volcanism Program/Smithsonian Institution, viewed 4 December 2020, <https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-05-12/readme.md>

## Modelling frequency by crust type and frequency by tectonic setting
One of the initial goals was to fit a linear model to frequency by crust type and tectonic setting separately. When carrying this out the values given by the output were plausible when viewing the graphs and a small amount of prior knowledge however the evalutations showed that linear models were not appropriate. The adjusted r squared values for both sets of data were very small showing the models were only accurate for a very small proportion of the data, in addition to this both the residuals plots showed residuals in distinct clusters that, when coloured by crust or tectonic settings, showed they were grouped by the categories of variable.
At this point we considered exploring other models however we decided that the use bar charts and other visualizations to explain the trends in frequency, tectonic settings and crust types rather than trying to fit a model that may not cover such a wide range of data efficiently.

##Number of Volcanos by Crust type.
The bar plot shows the number of eruptions that have been recorded on each separate crust type. The unknown volcanos where filtered out of the data to try to get a clearer idea of any relationship displayed. Its clear from the graph that the majority of eruptions occur on continental plates, being over 4 times greater than those recorded on oceanic plates and about 10 times greater than those recorded on intermediate plates. 
The ambiguity of an intermediate plate may explain the small number of observations there. It is also possible that the lower number of eruptions on Oceanic plates is down to more remote locations of the volcanoes. The data set goes back to very early years so it's plausible that eruptions recorded at these times would only have been those where a town or village was impacted by the volcano directly. Eruptions in the middle of the ocean likely went unrecorded as there was no way of knowing they were occurring.


## Number of Eruptions by Country and Volcano Type
Plot shows the frequency of eruptions in each country, grouped by volcano type. We choose to only show the top 25 groups of country and volcano types to avoid overwhelming so this is only a sample of the data. It can be clearly seen that Japan has the highest number of eruptions, mainly from strato, caldera and complex volcanoes, closely followed by Indonesia, and the composition of the eruptions by volcano type is similar.The US also has a high number of volcanic eruptions, with eruptions almost evenly split between strato and shield volcanoes. If we look at the countries of South America; Mexico,Equador,Costa Rica and Chile, they all have a relatively similar number of eruptions, solely composed of strato volcanoes. This backs up our conclusion that tectonic plate type, and crust type affect the number of eruptions, and also suggests that these factor also affect the volcano type. If we had more time I think it would be an interesting extension of our project to investigate the relationship between the factors which affect the number of volcano eruptions. The majority of this plot if red which indicates majority of volcanoes with high number of eruptions in our data set are stratovolcanoes. The variation in number of eruptions as primary volcano types varies, may lead us to believe that primary volcano type is a factor which affects the number of eruptions. 

# Number of Eruptions by Volcano Type
This can be seen more clearly in this bar plot which shows the number of eruptions grouped by primary volcano type, in particular it shows 9 primary volcano types with the top 9 number of eruptions. This very clearly shows that stratovolcanoes are responsible for the greatest number of eruptions, with nearly 6000 eruptions in the dataset being eruptions of stratovolcanoes. This is significantly more eruptions than shield and Caldera volcanoes, which have the second and third highest number of eruptions, with around 1000 eruptions recorded in the dataset. 
We could conclude that volcano type is a factor which affects the number of eruptions, but I believe to support this conclusion it is important to take into the consideration the number of each volcano type in the dataset.

## Number of Volcanoes by Volcano Type
This bar plot shows the number of volcanoes grouped by primary volcano type. It shows there are 23 different primary volcano types recorded in the dataset. Stratovolcanoes are the most abundant volcano types, therefore it makes sense that they cause the most eruptions, however there are approximately 450 stratovolcanoes, which are responsible for 6000 eruptions in the dataset. Therefore Stratovolcanoes must erupt several times within there lifetime, approximately 13 times each if we were to take a raw average.
You can also see if we compare the bar plot of Number of eruptions by volcano type and the bar plot for number of volcanoes by volcano type, that the plot take a similar shape, which may be an indicator that the number of eruptions of each volcano type is proportional to the number of volcanoes of that type.


## References

- https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-05-12/readme.md
- https://rstudio.github.io/leaflet/markers.html
- https://arilamstein.com/blog/2016/03/21/mapping-election-results-r-choroplethr/
- https://www.britannica.com/science/volcano
- 

