---
title: "Project proposal"
author: "x404-pagenotfound"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(janitor)
```

```{r read-in-data, message=FALSE}
volcano <- read_csv("data/volcano.csv")
eruptions <- read_csv("data/eruptions.csv")
events <- read_csv("data/events.csv")

```



## 1. Introduction
## The introduction should introduce your general research question and your data (where it came from, how it was collected, what are the cases, what are the variables, etc.).
As a team we choose to use the Volcano Eruptions data set from the tidytuesday package for our package, due to the interesting nature of the data, and ....
--large data set therefore can answer many data analytical questions, variety of variables
-- large number of both observations and variables
--mix of both continuous and categorical variables

The data in the Volcano eruption dataset comes from the Smithsonian Institution's Global Volcanism Program (GVP), which reports on the world's volcanos, and their eruptions, both current and dating back to 10,000 years ago. The data was collected from the Volcanos of the World database which is maintained by the GVP. The GVP aim "document, understand, and disseminate information about global volcanic activity," as their website https://volcano.si.edu/ states.

## 2. Data
## Place your data in the /data folder, and add dimensions and codebook to the README in that folder. Then print out the output of glimpse() or skim() of your data frame.


## Need to move data folder back into main project folder

```{r glimpse-data}

glimpse(volcano)
glimpse(eruptions)
glimpse(events)

```



## 3. Data analysis plan



