---
title: "Project proposal"
author: "x404-pagenotfound"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(janitor)
library(glue)
```

```{r read-in-data, message=FALSE}
volcano <- read_csv("data/volcano.csv")
eruptions <- read_csv("data/eruptions.csv")
events <- read_csv("data/events.csv")

```



## 1. Introduction
## The introduction should introduce your general research question and your data (where it came from, how it was collected, what are the cases, what are the variables, etc.).
As a team we choose to use the Volcano Eruptions data set from the tidytuesday package for our package, due to the interesting nature of the data, the number of observations, and the variety and number of variables. 

The data in the Volcano eruption dataset comes from the Smithsonian Institution's Global Volcanism Program (GVP), which reports on the world's volcanos, and their eruptions, both current and dating back to 10,000 years ago. The data was collected from the Volcanos of the World database which is maintained by the GVP. The GVP's aim is to "document, understand, and disseminate information about global volcanic activity," as their website https://volcano.si.edu/ states.
Each observation in the dataset volcano contains information on one of the earth's volcanoes, information such as the type of the volcano, where it is located and the types of rocks that make up te volcano. In the eruptions dataset each observation documents an eruption of one of the earth's volcano, documenting where and when each eruption occurs, as well as the eruption type and category.

## 2. Data


```{r glimpse-data}

glimpse(volcano)
glimpse(eruptions)
glimpse(events)

```


## 3. Data analysis plan

```{r create-start-end-date}
eruptions %>%
  mutate(
    start_date = glue("{start_year}{start_month}{start_day}"),
    end_date = glue("{end_year}{end_month}{end_day}")
    )
```
##library(lubridate)
##with(df1, ymd_h(paste(year, month, day, hour, sep= ' ')))

Section 3 - Data analysis plan:

The outcome (response, Y) and predictor (explanatory, X) variables you will use to answer your question.
The comparison groups you will use, if applicable.
Very preliminary exploratory data analysis, including some summary statistics and visualizations, along with some explanation on how they help you learn more about your data. (You can add to these later as you work on your project.)
The statistical method(s) that you believe will be useful in answering your question(s). (You can update these later as you work on your project.)
What results from these specific statistical methods are needed to support your hypothesized answer? 
--avg length eruption
--avg length time between eruptions
--which volcanos erupt most often --> by tectonic plate



