---
title: "Model frequency~techtonics"
author: "Bridget"
date: "11/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages
```{r load-packages}
library(tidyverse)
library(tidymodels)
library(janitor)
library(glue)
library(choroplethr)
library(choroplethrMaps)
library(leaflet)
```

## Read in Data

```{r read-in-data, message=FALSE}
setwd("/cloud/project/proposal")
volcano <- read_csv("../data/volcano.csv")
eruptions <- read_csv("../data/eruptions.csv")
events <- read_csv("../data/events.csv")

```

## Tidy Data Needed
```{r seperate-techtonics}
volcano_tecs <- volcano%>%
  #filter out Vitim Volcanic Field volcano as tec settings are unknown
  filter(tectonic_settings != "Unknown")%>%
  #seperate the tec setting from the crust 
  separate(tectonic_settings, c("techtonic_setting", "crust"),
                    sep = "([/])")
  
```

```{r eruption-frequency}
eruption_tecs<- left_join(volcano_tecs, eruptions,
                          by=c("volcano_number", "volcano_name",
                               "latitude", "longitude"))
  
eruption_tecs_fqncy <- eruption_tecs%>%
  count(volcano_number, volcano_name, techtonic_setting, crust) %>% 
  mutate(frequency = as.factor(n))%>%
  #str_trim("crust", side = "left")%>%
  #ideally remove whitespace for easier use but leave for later
  filter(crust != " Crustal thickness unknown")
  

  
```

## Initial Visualisations
```{r viz}
eruption_tecs_fqncy%>%
  group_by(crust)%>%
  ggplot(aes(y = fct_rev(fct_infreq(crust)), fill = crust))+
  geom_bar()+
  labs(y = "Crust Type (thickness)",
       x = "Eruptions", 
       title = "Frequency of eruptions on different crusts")+
  scale_fill_manual(values = c(
    " Continental crust (>25 km)" = "#d73027",
    " Oceanic crust (< 15 km)" = "#4575b4",
    " Intermediate crust (15-25 km)" = "#fee090"))+
  guides(fill = FALSE)
```  

```{r viz2}
eruption_tecs_fqncy%>%
  group_by(techtonic_setting)%>%
  ggplot(aes(y = techtonic_setting, fill = techtonic_setting))+
  geom_bar()+
    labs(y = "Tectonic Setting",
       x = "Eruptions", 
       title = "Frequency of eruptions for different tectonic settings")+
  scale_fill_manual(values = c(
    "Intraplate " = "#a50026",
    "Rift zone " = "#313695",
    "Subduction zone " = "#4d9221"))+
  guides(fill = FALSE)
  

```
The first of the two visualizations above show that the majority of erruptions in the data set occur on continental crust which is over 25km thick.
The second shows most erruptions in the dataset have occured from volcanos with the techtonic setting in the subduction zone

I aim to see if there is a model to support the hypothesis that it is a there is more likely to be a volcanic erruption if the volcano is lying on continental crust and has techtonic settings in the subduction zone than other crust types and techtonic settings.

## Linear Models
```{r tec_setting-model}
tec_fit <- linear_reg()%>%
  set_engine("lm")%>%
  fit(n ~ techtonic_setting, data= eruption_tecs_fqncy)

  glance(tec_fit)$adj.r.squared

```
The adjusted R squared for the linear model shows it only accounts for 1.2% of the data and therefore a linear model isn't appropriate for the data, I will continue to explore general models to fit the data better.

```{r tec_setting-residuals}
tec_fit_aug <- augment(tec_fit$fit)

ggplot(tec_fit_aug, mapping=aes( x=.fitted, y =.resid, 
                                 colour = techtonic_setting)) +
	geom_jitter(alpha=0.8) +
	geom_hline(yintercept=0, color="gray", lty="dashed" )+
	labs(x="Predicted Frequency of Volcanic Eruptions", y="Residuals",
	     colour = "Tectonic Setting")+
    scale_colour_manual(values = c(
    "Intraplate " = "#a50026",
    "Rift zone " = "#313695",
    "Subduction zone " = "#4d9221"))
```

The residuals of predicted frequency seem to be in 3 distinct clusters. By adding an aesthetics arguement for colour of each tectonic setting it is clear that each of these clusters represents each setting. The plot further supports that a linear model is not appropriate for the data in question.

```{r crust-model}
crust_fit <- linear_reg()%>%
  set_engine("lm")%>%
  fit(n ~ crust, data= eruption_tecs_fqncy)

  glance(crust_fit)$adj.r.squared

```
The adjusted R squared for the linear model shows it only accounts for 0.2% of the data and therefore a linear model isn't appropriate for the data, I will continue to explore general models to fit the data better.

```{r crust-fit-residuals}
crust_fit_aug <- augment(crust_fit$fit)

ggplot(crust_fit_aug, mapping=aes( x=.fitted, y =.resid, colour = crust)) +
	geom_jitter(alpha=0.8) +
	geom_hline(yintercept=0, color="gray", lty="dashed" )+ 
	labs(x="Predicted Frequency of Volcanic Eruptions", y="Residuals")+
  scale_colour_manual(values = c(
    " Continental crust (>25 km)" = "#d73027",
    " Oceanic crust (< 15 km)" = "#4575b4",
    " Intermediate crust (15-25 km)" = "#fee090"))
```
The residuals of predicted frequency seem to be in 3 distinct clusters. By adding an aesthetics arguement for colour of each crust type and thickness it is clear that each of these clusters represents each crust type. The plot further supports that a linear model is not appropriate for the data in question.

## Split Data
```{r split-data}
set.seed(321)
tec_split <- initial_split(eruption_tecs_fqncy, prop = 0.80)
train_data_tec <- training(tec_split)
test_data_tec  <- testing(tec_split)
```

## General Model
```{r general-model-tecs}
#tec_fit_log <-
  logistic_reg()%>%
  set_engine("glm")%>%
  fit(frequency ~ techtonic_setting, data= test_data_tec)%>%
  tidy()

```

```{r general-model-crust}
#crust_fit_log <- 
  logistic_reg()%>%
  set_engine("glm")%>%
  fit(frequency ~ crust, data= test_data_tec)%>%
  tidy()
```
## Map visualizations
```{r tec-map}
tec_pal<- colorFactor(
  palette = c("#a50026", "#313695", "#4d9221"),
  domain = eruption_tecs$techtonic_setting
)
#red=intraplate
#blue=riftzone
#green=subductionzone


leaflet(eruption_tecs) %>% 
  addTiles() %>%
  addCircleMarkers(
    ~longitude, ~latitude,
    radius = 3,
    color = ~tec_pal(techtonic_setting),
    stroke = FALSE, 
    fillOpacity = 0.4
  )
```

```{r crust-map}
crust_pal<- colorFactor(
  palette = c("#d73027", "#762a83", "#fee090", "#4575b4"),
  domain = eruption_tecs$crust
)
#red=continental
#yellow=intermediate
#blue=oceanic
#purple=unknown


leaflet(eruption_tecs) %>% 
  addTiles() %>%
  addCircleMarkers(
    ~longitude, ~latitude,
    radius = 3,
    color = ~crust_pal(crust),
    stroke = FALSE, 
    fillOpacity = 0.4
  )
```
























