---
title: "Factors affecting frequency of Volcanic eruptions"
subtitle: "And Visualisations of their Global Distrubution"
author: "x404-pagenotfound <br> Emma, Bridget, Sachin"
institute: "University of Edinburgh"
date: "4 December 2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(broom)
library(janitor)
library(glue)
#for choropleth map
library(choroplethr)
library(choroplethrMaps)
#for leaflet interactive maps, and layering maps
library(leaflet)
library(rgdal)

```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r load-data, include=FALSE}
# Load your data here
setwd("/cloud/project/proposal")
volcano <- read_csv("../data/volcano.csv") #contains info on every volcano
eruptions <- read_csv("../data/eruptions.csv") #contains info about every eruptions
events <- read_csv("../data/events.csv")
```

```{r data-for-map, include=FALSE}
#Join eruptions and volcano to match each eruption to its corresponding volcano
volcanic_eruption <- left_join(volcano, eruptions, by="volcano_number") #match each eruption with its corresponding volcano in the volcano df
#load df with list of countries
data("country.regions")
#create CountVolc which stores number of eruptions for each country
CountVolc = group_by(volcanic_eruption, country) %>% summarise(value = n())
CountVolc = as.data.frame(CountVolc)
CountVolc[,1] = tolower(CountVolc[,1])
rownames(CountVolc) = CountVolc$country

#some eruptions are on the border between two countries, so we split that count between each country
CountVolc["chile","value"] = CountVolc["chile","value"] + CountVolc["chile-argentina","value"]/2
CountVolc["argentina","value"] = CountVolc["argentina","value"] + CountVolc["chile-argentina","value"]/2

#some country names must be changed to get the count
CountVolc$country=gsub("united states", "united states of america", CountVolc$country)

#Aggregating the counts into unique country names
CountVolcUni <- group_by(CountVolc, country) %>%
  summarise(value = sum(value))
CountVolcUni = as.data.frame(CountVolcUni)
rownames(CountVolcUni) = CountVolcUni$country

```

class: center, middle

## What factors affect the frequency of volcanic eruptions?


---


background-image: url(https://camo.githubusercontent.com/4295e3eef5545f83277639df7e7b73fa7de3bc5722c0da60d5ff44b6e1f3df8d/68747470733a2f2f75706c6f61642e77696b696d656469612e6f72672f77696b6970656469612f636f6d6d6f6e732f7468756d622f332f33342f4572757063692543332542336e5f656e5f656c5f766f6c632543332541316e5f536162616e636179612532435f5065722543332542412e6a70672f3132383070782d4572757063692543332542336e5f656e5f656c5f766f6c632543332541316e5f536162616e636179612532435f5065722543332542412e6a7067)

class: inverse, center, middle

# Volcano Data Set

---
# World Map showing Number of Volcanic Eruptions

```{r World-Map, echo=FALSE, warning=FALSE, message=FALSE}
#create tibble with rownames as countries
CountVolcFake = country.regions
CountVolcFake$value = 0
rownames(CountVolcFake) = CountVolcFake$region
#choose the countries in both tibbles
ii=intersect(CountVolcUni$country,country.regions$region)
CountVolcFake[ii, "value"] = (CountVolcUni[ii, "value"])
#create choropleth map with palette
country_choropleth(CountVolcFake, num_colors = 9) +
scale_fill_brewer(palette="YlOrRd") +
labs(title = "Number of Volcanic Eruptions by Country")




#find countries that aren't included
MissingCountries = CountVolcUni$country[!CountVolcUni$country %in% country.regions$region]
```


---

```{r, echo=FALSE}
#clustered leaflet interactive map
set.seed(1)
ran_rows <- sample(nrow(volcanic_eruption), 1000) #random sample for testing
leaflet(data = volcanic_eruption) %>% 
  addTiles() %>%
  addMarkers(~longitude.x, ~latitude.x, popup = ~as.character(volcano_name.x), label = ~as.character(volcano_name.x), clusterOptions = markerClusterOptions()) %>%
  setView(3, -5, zoom = 2.5) #set starting view and zoom
```

---
```{r, include=FALSE}
my_spdf <- readOGR("/cloud/project/presentation/presentation_files/tectonicplates-master/PB2002_plates.shp") #tectonic plate shape data (.shp)
```

```{r, echo=FALSE}
beatCol <- colorFactor(palette = 'RdYlGn', factor(volcanic_eruption$vei)) #random color palette for plotting by magnitude (vei)

leaflet(volcanic_eruption) %>%
  addTiles() %>% # get map tiles from google maps
  addCircleMarkers( #add points
    ~longitude.x, ~latitude.x,
    radius = ~sqrt(volcanic_eruption$vei*5), #radius of points depends on eruption magnitude
    color = "red",#~beatCol(factor(volcanic_eruption$vei)) will be used if we want to color by magnitude
    stroke = FALSE, fillOpacity = 0.4
  ) %>%
  addPolylines(data = my_spdf, weight = 2) %>% #add tectonic boundaries
  setView(3, -5, zoom = 2.5) #set starting view and zoom
```

```{r Proof-that-magnitude-isnt-correlated-to-position, include=FALSE}
leaflet(subset(volcanic_eruption, vei ==2)) %>% #vei == 1,2,3,4,5,6,7 to show a map for each magnitude
  addTiles() %>%
  addCircleMarkers(
    ~longitude.x, ~latitude.x,
    radius = 3,
    color = "red",#~beatCol(factor(volcanic_eruption$vei)),
    stroke = FALSE, fillOpacity = 0.4
  ) %>% 
  addPolylines(data = my_spdf, weight = 2) %>%
  setView(-1.479538, 53.380851, zoom = 3)
```


---

#What are the variables crust and tectonic setting?
Volcano formation concerns the top two layers of the earth- the crust(composed of sections known as tectonic plates, this is the surface) and the mantle(the second outermost layer formed of partially molten rock (magma) which moves very slowly).

.pull-left[
#Plates and Crust type
The crust variable refers to the crust type and thickness of the plate the volcano is on,
In our data there are four categories:

-Continental (>25 km)- land plates 

-Oceanic (< 15 km)- sea plates

-Intermediate (15-25 km) - unclear but potentially plates that lie outside the average thickness for oceanic or continental 

-Unknown 
]

.pull-right[

#Tectonic Settings
The tectonic setting is the movement of the plates and mantle that causes the volcano to form, given as the variables:

*Subduction zone:* 
An oceanic plate moves beneath a continental plate, forming more magma which then forms a volcano on the continental plate

*Rift zone:* 
Two plates move away from each other forming a gap for mantle to move through forming a volcano.

*Intraplate:* 
A much hotter area of the mantle causes magma to rise to the surface to form a volcano.]
---

# Map eruptions by tectonic plate type
.pull-left[
```{r seperate-techtonics, echo=FALSE}
volcano_tecs <- volcano%>%
  #filter out Vitim Volcanic Field volcano as tec settings are unknown
  filter(tectonic_settings != "Unknown")%>%
  #seperate the tec setting from the crust 
  separate(tectonic_settings, c("techtonic_setting", "crust"),
                    sep = "([/])")
  
```

```{r eruption-frequency, echo=FALSE}
eruption_tecs<- left_join(volcano_tecs, eruptions,
                          by=c("volcano_number", "volcano_name",
                               "latitude", "longitude"))
  
eruption_tecs_fqncy <- eruption_tecs%>%
  count(volcano_number, volcano_name, techtonic_setting, crust) %>% 
  mutate(frequency = as.factor(n))%>%
  #to try to fit best model remove unknown crusts
  filter(crust != " Crustal thickness unknown")
  

  
```

```{r tec-map, echo = FALSE}
#set colours for each tectonic setting
tec_pal<- colorFactor(
  palette = c("#a50026", "#313695", "#4d9221"),
  domain = eruption_tecs$techtonic_setting
)
#red=intraplate
#blue=riftzone
#green=subductionzone


leaflet(eruption_tecs) %>% 
  addTiles() %>%
  addCircleMarkers(
    ~longitude, ~latitude,
    radius = 3,
    #add colour argument dependent on techtonic_setting
    color = ~tec_pal(techtonic_setting),
    stroke = FALSE, 
    fillOpacity = 0.4
  ) %>%
  setView(3, -5, zoom = 1.5) #set starting view and zoom
```
]
.pull-right[
#Tectonic Setting
red= Intraplate

blue= Rift Zone

green= Subduction Zone
]

---

# Map of Crust type
.pull-left[
```{r crust-map, echo=FALSE}
#set colours for each crust type
crust_pal<- colorFactor(
  palette = c("#d73027", "#762a83", "#fee090", "#4575b4"),
  domain = eruption_tecs$crust
)
#red=continental
#yellow=intermediate
#blue=oceanic
#purple=unknown


leaflet(eruption_tecs) %>% 
  addTiles() %>%
  addCircleMarkers(
    ~longitude, ~latitude,
    radius = 3,
    #add colour arguement dependant on crust
    color = ~crust_pal(crust),
    stroke = FALSE, 
    fillOpacity = 0.4
  ) %>%
  setView(3, -5, zoom = 1.5) #set starting view and zoom
```
]
.pull-right[

#Crust types
red=continental

yellow=intermediate

blue=oceanic

purple=unknown
]
---

# Model frequency by crust type

A linear model was initially fit to the data to try to predict how many times the volcano would erupt depending on the crust type, however:

.pull-left[
#Residuals
The residuals plot shows three distinct clusters of data for each different crust type:
```{r crust-model, out.width = "50%", echo= FALSE}
crust_fit <- 
linear_reg()%>%
  set_engine("lm")%>%
  fit(n ~ crust, data= eruption_tecs_fqncy)

```
```{r crust-fit-residuals, echo= FALSE}
crust_fit_aug <- augment(crust_fit$fit)

ggplot(crust_fit_aug, mapping=aes( x=.fitted, y =.resid,
                                   #colour added to investigate clusters of variables
                                   colour = crust)) +
	geom_jitter(alpha=0.8) +
	geom_hline(yintercept=0, color="gray", lty="dashed" )+ 
	labs(x="Predicted Frequency of Volcanic Eruptions", y="Residuals", colour = "Crust type (thickness)")+
  #Set same colours as shown on maps
  scale_colour_manual(values = c(
    " Continental crust (>25 km)" = "#d73027",
    " Oceanic crust (< 15 km)" = "#4575b4",
    " Intermediate crust (15-25 km)" = "#fee090"))
```
]

.pull-right[
#Adjusted R Squared

In addition, the adjusted R squared value for this model was 0.002351574, meaning the model is only accurate for a very small proportion of the data.


#Conclusion
The two separate evaluations allowed us to conclude that the linear model was unsuitable for the data in hand.]

---
# Model frequency by tectonic setting

When modeling frequency by tectonic setting we encountered a very similar occurence.

.pull-left[
#Residuals

The residuals plot was again split into 3 clusters, one for each tectonic setting.

```{r tec_setting-model, echo=FALSE}
tec_fit <- linear_reg()%>%
  set_engine("lm")%>%
  fit(n ~ techtonic_setting, data= eruption_tecs_fqncy)
```
```{r tec_setting-residuals, echo=FALSE}
tec_fit <- linear_reg()%>%
  set_engine("lm")%>%
  fit(n ~ techtonic_setting, data= eruption_tecs_fqncy)


tec_fit_aug <- augment(tec_fit$fit)

ggplot(tec_fit_aug, mapping=aes( x=.fitted, y =.resid, 
                                 #colour added to investigate clusters of variables
                                 colour = techtonic_setting)) +
	geom_jitter(alpha=0.8) +
	geom_hline(yintercept=0, color="gray", lty="dashed" )+
	labs(x="Predicted Frequency of Volcanic Eruptions", y="Residuals",
	     colour = "Tectonic Setting")+
   #Set same colours as shown on maps
    scale_colour_manual(values = c(
    "Intraplate " = "#a50026",
    "Rift zone " = "#313695",
    "Subduction zone " = "#4d9221"))
```
]

.pull-right[
#Adjusted R squared
The adjusted R squared value was 0.01215089, again only accounting for a very small portion of the data.

# What we decided

-Linear models weren't appropriate

-Visualiations and comparison would answer our research question to the depth we'd initially planned for so general models wouldn't be necessary
]
---
# Eruption Frequency on each crust

```{r viz, echo=FALSE}
eruption_tecs_fqncy%>%
  group_by(crust)%>%
  ggplot(aes(# reorder for more direct comparison
    y = fct_rev(fct_infreq(crust)), fill = crust))+
  geom_bar()+
  labs(y = "Crust Type (thickness)",
       x = "Eruptions", 
       title = "Frequency of eruptions on different crusts")+
   #Set same colours as shown on maps
  scale_fill_manual(values = c(
    " Continental crust (>25 km)" = "#d73027",
    " Oceanic crust (< 15 km)" = "#4575b4",
    " Intermediate crust (15-25 km)" = "#fee090"))+
  #Legend not required
  guides(fill = FALSE)
```  

---

# Eruption frequency for each tectonic setting

```{r viz2, echo = FALSE}
eruption_tecs_fqncy%>%
  group_by(techtonic_setting)%>%
  ggplot(aes(y = techtonic_setting, fill = techtonic_setting))+
  geom_bar()+
    labs(y = "Tectonic Setting",
       x = "Eruptions", 
       title = "Frequency of eruptions for different tectonic settings")+
   #Set same colours as shown on maps
  scale_fill_manual(values = c(
    "Intraplate " = "#a50026",
    "Rift zone " = "#313695",
    "Subduction zone " = "#4d9221"))+
  #Legend not required
  guides(fill = FALSE)
  

```


---

# Bar chart frequency by country by volcano type

```{r eruptions-by-country-and-volcano-type, echo=FALSE}

volcanic_eruption %>%
  group_by(country, primary_volcano_type) %>%
   mutate( freq_eruptions=n(),
          primary_volcano_type=if_else(primary_volcano_type=="Stratovolcano(es)", "Stratovolcano", primary_volcano_type) ) %>%
  arrange(desc(freq_eruptions)) %>%
  distinct(country, primary_volcano_type, freq_eruptions) %>%
  head(25)%>%
  ggplot(aes(x=country,y=freq_eruptions)) +
  geom_col(aes(fill=primary_volcano_type)) +
  coord_flip() +
  labs(x="Country",y="Number of Eruptions", title="Frequency of Eruptions by Country", subtitle="and Vocano Type", fill="Primary Volcano Type") +
  scale_fill_brewer(palette="YlOrRd")


```
---
# Bar chart frequency by volcano type

```{r eruptions-by-volcano-type, echo=FALSE}

volcanic_eruption %>%
  group_by(primary_volcano_type) %>%
  mutate( freq_eruptions=n(),
          primary_volcano_type=if_else(primary_volcano_type=="Stratovolcano(es)", "Stratovolcano", primary_volcano_type) ) %>%
  arrange(desc(freq_eruptions)) %>%
  distinct(primary_volcano_type, freq_eruptions) %>%
  head(10)%>%
  ggplot(aes(x=fct_rev(as.factor(primary_volcano_type)),y=freq_eruptions,fill=primary_volcano_type)) +
  geom_col() +
  coord_flip() +
  labs(x="Primary Volcano Type",y="Number of Eruptions", title="Frequency of Eruptions by Vocano Type") +
  scale_fill_brewer(palette="YlOrRd")

```
---
```{r count-volcanoes-by-volcano-type, echo=FALSE}

volcano %>%
  group_by(primary_volcano_type) %>%
  mutate( primary_volcano_type=case_when(
    primary_volcano_type=="Stratovolcano(es)"~"Stratovolcano",
    primary_volcano_type=="Shield(s)"~"Shield",
    primary_volcano_type=="Pyroclastic cone(s)"~"Pyroclastic cone",
    TRUE ~ primary_volcano_type
    )) %>%
  ggplot(aes(x=fct_rev(as.factor(primary_volcano_type)),fill=primary_volcano_type)) +
  geom_bar() +
  coord_flip() +
  labs(x="Primary Volcano Type",y="Count", title="Number of Volcanoes by Vocano Type")

```

---

# Conclusion
We believe the factors which affect volcanic eruptions according to our investigation are : 
-Location of volcano (position on tectonic plate)
-Tectonic Plate Setting
-Tectonic Plate Type
-Crust Type
-Primary Volcano Type

We also hypothesize that there may be a relationship between these factors, based on our findings.
---
#Future research/Investigation
- Investigate the relationship between the factors which affect the volcanic eruptions - cause them to arise, how frequently they arise etc